Ecosystem functional response across precipitation extremes in a sagebrush steppe

### 2017-09-18 manuscript/usses_water_ms.Rmd
adds pba edits

---
title: Ecosystem functional response across precipitation extremes in a sagebrush steppe
preprint: false
author: 
  - name: Andrew T. Tredennick
    affiliation: 1
    corresponding: true
    email: atredenn@gmail.com
  - name: Andrew R. Kleinhesselink
    affiliation: 2
  - name: Bret Taylor
    affiliation: 3
  - name: Peter B. Adler
    affiliation: 1
affiliation:
  - code: 1
    address: Department of Wildland Resources and the Ecology Center, Utah State University, Logan, Utah 84322
  - code: 2
    address: Department of Ecology and Evolutionary Biology, University of California, Los Angeles, Los Angeles, California 90095
  - code: 3
    address: United States Department of Agriculture, Agriculture Research Station, U.S. Sheep Experiment Station, Dubois, Idaho 83423
abstract: |
  **Background.** Precipitation is predicted to become more variable in the western United States, meaning years of above and below average precipitation will become more common. Periods of extreme precipitation are major drivers of ecosystem functioning in water limited grasslands, but how ecosystems respond to precipitation may change as the duration of above and below average periods increases. Changes in ecosystem functional response could reflect compensatory changes in species composition or species reaching physiological thresholds at extreme precipitation levels.

  **Methods.** We conducted a five year precipitation manipulation experiment in a sagebrush steppe ecosystem in Idaho, United States. We used drought and irrigation treatments (approximately 50% decrease/increase) to investigate whether ecosystem functional response remains consistent at extreme precipitation levels. We recorded data on aboveground net primary productivity (ANPP), species abundance, and soil moisture. We fit a generalized linear mixed effects model to determine if the relationship between ANPP and soil moisture differed among treatments. We used nonmetric multidimensional scaling to quantify community composition over the five years.

  **Results.** Ecosystem functional response, defined as the relationship between soil moisture and ANPP was similar among drought and control treatments, but the irrigation treatment had a lower slope than the control treatment. However, ANPP response to availabile soil moisture was weak and uncertain regardless of treatment, with all slopes overlapping zero. There was also large variation in ANPP within-years. Plant community composition was remarkably stable over the course of the experiment and did not differ among treatments.

  **Discussion.** Despite some evidence that ecosystem functional response was less positive under extreme wet conditions, the response of ANPP to soil moisture was consistently weak and community composition was remarkably stable. Ecosystem functional responses across treatments was not related to compensatory shifts at the plant community level, but instead reflects the weak and variable responses of individual species. Such responses may be due to bet hedging strategies where species respond weakly to even extreme levels of soil moisture to ensure low variance of long-term success.
header-includes: >
  \usepackage{setspace}
  \usepackage{todonotes}
  \usepackage{rotating}
  \usepackage{color, soul}
  \usepackage{sectsty}
  \usepackage{bm,mathrsfs}
bibliography: ~./usses_water.bib
output: 
  rticles::peerj_article:
    number_sections: false
    base_format: rmarkdown::pdf_document # bookdown::pdf_document2 # for using \@ref()
---

@@

The functional response of aboveground net primary productivity (ANPP) to water availability (e.g., soil moisture) can be characterized by fitting a model to historical observations of ANPP and soil moisture at a given site.
However, the fitted functional response may provide an incomplete picture because future conditions are likely to be outside the historical range of variability [@Smith2011].
For example, historical trends may underestimate the potential for the soil moisture-ANPP relationship to saturate or weaken if soil moisture is pushed far beyond typical levels.
Saturating relationships are actually common [@Hsu2012; @Gherardi2015a], perhaps because other resources, like nitrogen, become more limiting in wet years than dry years.
Knowing the curvature of the soil moisture-ANPP relationship at extreme precipitation levels is critical for understanding how ecosystems will respond to chronic alterations in water availability.

Another problem with relying on historical ecosystem functional responses is that they are not static.
Changes in species identities and abundances can alter an ecosystem's functional response to water availability because different species have different physiological thresholds for producing biomass.
@Smith2009 introduced the 'Hierarchical Response Framework' for understanding the interplay of community composition and ecosystem functioning in response to resource manipulations over time.
In the near term, ecosystem functioning such as ANPP will reflect the physiological responses of individual species to the manipulated resource level.
For example, ANPP may decline under simulated drought because the initial community consisted of drought-intolerant species [@Hoover2014].
Ecosystem functioning may recover over longer time spans as new species colonize or initial species reorder in relative abundance.
For example, ANPP may initially decline, but eventually rise back to pre-treatment levels once drought-tolerant species become more abundant and compensate for drought-intolerant species [@Hoover2014].
It is also possible that ecosystem functioning shifts to a new mean state, reflecting the suite of species in the new community [@Knapp2012].

Manipulating potentially limiting resources, like precipitation, offers a route to understanding how ecosystems will respond to resource levels that fall outside the historical range of variability [@Avolio2015; @Gherardi2015; @Knapp2017].
Altering the amount of precipitation over many years should provide insight into the time scales at which water-limited ecosystems respond to chronic resource alteration.
We propose four alternative predictions for the effect of precipitation manipulation on the ecosystem functional response to soil moisture based on the Hierarchical Response Framework (Fig. 1).
We define 'ecosystem functional response' as the relationship between available soil moisture and ANPP.
The four predictions are based on possible outcomes at the community (e.g., community composition) and ecosystem (e.g., soil moisture-ANPP regression) levels.

First, altered precipitation changes neither ecosystem functional response nor community composition (Fig. 1, top left).
In this case, changes in ANPP simply follow the soil moisture-ANPP relationship under ambient conditions.
This corresponds to the early phases of the Hierarchical Response Framework, where ecosystem response follows the physiological responses of individual species.
Second, the ecosystem functional response changes but community composition remains the same (Fig. 1, top right).
A saturating soil moisture-ANPP response fits this scenario, where individual species hit physiological thresholds or are limited by some other resource.
Third, the ecosystem functional response is consistent but underlying community composition changes (Fig. 1, bottom left).
In this case, changes in species' identities or abundances occur in response to altered precipitation levels and species more suited to the new conditions compensate for reduced function of initial species.
Fourth, and last, both ecosystem functional response and community composition change (Fig. 1, bottom right).
New species, or newly abundant species, with different physiological responses completely reshape the ecosystem functional response.

All four outcomes are possible in any given ecosystem, but the time scales at which the different scenarios play out likely differ [@Smith2009; @Wilcox2016; @Knapp2017].
Thus, our task is not to test the validity of the Hierarchical Response Framework, but rather to amass information on how quickly ecosystem functional responses change in different ecosystems.
We also need to understand whether changes at the ecosystem level are driven by community level changes or individual level responses.

To that end, here we report the results of a five-year precipitation manipulation experiment in a sagebrush steppe grassland.
We imposed drought and irrigation treatments (approximately $\pm50\%$) and measured ecosystem (ANPP) and community (species composition) responses.
We focus on how the drought and irrigation treatments affect the relationship between available soil moisture and ANPP, and if community dynamics underlie the ecosystem responses.
In particular, we are interested in the consistency of the soil moisture-ANPP relationship among treatments.
Is the relationship steeper under the drought treatment, at low soil moisture?
Does the relationship saturate under the irrigation treatment, at high soil moisture?
To answer these questions we fit a generalized linear mixed effects model to test whether the regressions differed among treatments.
We also analyzed community composition over time, allowing us to place our experimental results within the framework our competing predictions (Fig. 1).

@@

## Study Area
We conducted our precipitation manipulation experiment in a sagebrush steppe community at the United States Sheep Experimental Station (USSES) near Dubois, Idaho (44.2$^{\circ}$ N, 112.1$^{\circ}$ W), 1500 m above sea level.
The plant community is dominated by the shrub *Artemesia tripartita* and three perennial bunchgrasses, *Pseudoroegneria spicata*, *Poa secunda*, and *Hesperostipa comata*.
During the period of our experiment (2011 -- 2015), average mean annual precipitation was 265 mm year$\phantom{}^{-1}$ and mean monthly temperature ranged from -5.2$^{\circ}$C in January to 21.8$^{\circ}$C in July.
Between 1926 and 1932, range scientists at the USSES established 26 permanent 1 m$^2$ quadrats to track vegetation change over time.
In 2007, we (well, one of us [P. Adler]) relocated 14 of the original quadrats, six of which were inside a large, permanent livestock exclosure.
We use these six plots as control plots that have received no treatment, just ambient precipitation, in the experiment described below.

## Precipitation Experiment
In spring 2011, we (well, two of us [A. Kleinhesselink and P. Adler]) established 16 new 1 m$^2$ plots located in the same exclosure as the six control plots.
We avoided areas on steep hill slopes, areas with greater than 20\% cover of bare rock, and areas with greater than 10\% cover of the shrubs *Purshia tridentata* and/or *Amelanchier utahensis*.
We established the new plots in pairs and randomly assigned each plot in a pair to receive a "drought" or "irrigation" treatment.

Drought and irrigation treatments were designed to decrease and increase the amount of ambient precipitation by 50\%.
To achieve this, we used a system of rain-out shelters and automatic irrigation [@Gherardi2013].
The rain-out shelters consisted of transparent acrylic shingles 1-1.5 m above the ground that covered an area of $2.5\times2$ m.
The shingles intercepted approximately 50\% of incoming rainfall, which was channeled into 75 liter containers.
Captured rainfall was then pumped out of the containers and sprayed on to the adjacent irrigation plot via two suspended sprinklers.
Pumping was triggered by float switches once water levels reached about 20 liters.
We disconnected the irrigation pumps each fall and reconnected them, often with difficulty, each spring.
The rain-out shelters remained in place throughout the year.

We monitored soil moisture in four of the drought-irrigation pairs using Decagon Devices (Pullman, Washington) 5TM and EC-5 soil moisture sensors.
We installed four sensors around the edges of each plot, two at 5 cm soil depth and two at 25 cm soil depth.
We also installed four sensors in areas nearby the four selected plot pairs to measure ambient soil moisture at the same depths.
Soil moisture measurements were automatically logged every four hours.
We coupled this temporally intensive soil moisture sampling with spatially extensive readings taken at six points within all 16 plots and associated ambient measurement areas.
These snapshot data were collected on 06-06-2012, 04-29-2015, 05-07-2015, 06-09-2015, and 05-10-2016\footnote{Dates formatted as: mm-dd-yyyy.} using a handheld EC-5 sensor.

Analyzing the response to experimental treatments was complicated by the fact that we did not directly monitor soil moisture in each plot on each day of the experiment.
Only a subset of plots were equipped with soil moisture sensors, and within those plots, one or more of the sensors frequently failed to collect data.
To remedy these problems, and to produce average daily soil moisture values for the ambient, drought, and irrigation conditions, we used a statistical model to predict the average treatment effects on soil moisture during the course of the experiment.

@@

We then modeled the daily deviation from ambient conditions of the drought and irrigation treatments using a linear mixed effects model with independent variables for treatment, season (winter, spring, summer, fall), rainfall, and all two-way interactions. 
Rainy days were defined as any day in which precipitation was recorded and average temperature was above 3$^{\circ}$C.
The day immediately following rainfall was also classified as rainy.
We fit the model using the ```lme4::lmer()``` function [@Bates2015] in R [@R2016], with random effects for plot group and date.
We weighted observations by the number of unique sensors or spot measurements that were taken in each plot on that day.
We then used the model to predict the average daily soil moisture in the treated plots based on the average daily ambient soil moisture.
We could only predict soil moisture in the treated plots on days for which we took at least one ambient soil moisture measurement.

## Data Collection
We estimated aboveground net primary productivity (ANPP) using a radiometer to relate ground reflectance to plant biomass [see @Byrne2011 for a review].
We recorded ground reflectance at four wavelengths, two associated with red reflectance (626 nm and 652 nm) and two associated with near-infrared reflectance (875 nm and 859 nm).
At each plot in each year, we took four readings of ground reflectances at the above wavelengths.
We also took readings in 12 (2015), 15 (2012, 2013, 2014), or 16 (2016) calibration plots adjacent to the experimental site, in which we harvested all aboveground biomass, dried it to a constant weight at 60$^{\circ}$C, and weighed it to estimate ANPP.

For each plot and year, we averaged the four readings for each wavelength and then calculated a greenness index based on the same bands used to calculate NDVI using the MODIS and AVHRR bands for NDVI.
We regressed the greenness index against the dry biomass weight from the ten calibration plots to convert the greenness index to ANPP.
We fit regressions to a MODIS-based index and an AVHRR-based index for each year and retained the regression with the better fit based on $R^2$ values.
We then predicted ANPP using the best regression equation for each year (Appendix 1).

Species composition data came from two sources: yearly census maps for each plot made using a pantograph [@Hill1920] and yearly counts of annual species in each plot.
We extracted the mapped information to obtain data on the density of all annuals and perennials forbs, the basal cover of perennial grasses, and the canopy cover of shrubs.
We made a large plot-treatment-year by species matrix, where columns were filled with either basal cover or density, depending on the measurement made for the particular species.
We standardized the values in each column so we could analyze the different types of data together.
This puts all abundance values on the same scale, meaning that common and rare species are weighted equally.
This approach is anti-conservative for detecting community change if we assume that rare species will respond more than common ones, which biases our approach toward detecting compositional changes.

@@

\noindent{}where $\mu_{i}$ is the deterministic prediction from the regression model for observation *i*, which is associated with plot *j* and treatment year *t*. 
$\boldsymbol{\beta}$ is the vector of coefficients for the fixed effects in the design matrix $\textbf{X}$. 
Each row of the design matrix represents a single observation ($\textbf{x}_i$) and is a vector with the following elements: 1 for the intercept, a binary 0 or 1 if the treatment is "drought", a binary 0 or 1 if the treatment is "irrigation", the scaled value of VWC, binary "drought" value times VWC, and binary "irrigation" value times VWC. 
Thus, our model treats "control" observations as the main treatment and then estimates intercept and slope offsets for the "drought" and "irrigation" treatments. 
In reference to our model, the hypotheses we wish to test are:

@@

@@

\noindent{}where $\Sigma$ is the variance-covariance matrix and *j(i)* reads as "plot *j* associated with observation *i*". 
The random year effects ($\boldsymbol{\eta}$) are drawn from a normal prior with mean 0 and standard deviation $\sigma_{\text{year}}$, which was drawn from a half-Cauchy distribution. 
A full description of our model is in Appendix 2.

@@

We used nonmetric multidimensional scaling (NMDS) based on Bray-Curtis distances to see if community composition differed among treatments through time.
We first calculated Bray-Curtis distances among all plots for each year of the experiment and then extracted those distances for use in the NMDS.
Some values of standardized species' abundances were negative, which is not allowed for calculating Bray-Curtis distances.
We simply added '2' to each abundance value to ensure all values were greater than zero.
We plotted the first two axes of NMDS scores to see if community composition overlapped, or not, among treatments in each year.
We used the ```vegan::metaMDS()``` function [@Oksanen2016] to calculate Bray-Curtis distances and then to run the NMDS analysis.
We used the ```vegan::adonis()``` function [@Oksanen2016] to perform permutational multivariate analysis of variance to test whether treatment plots formed distinct groupings.
To test whether treatment plots were equally dispersed, or not, we used the ```vegan::betadisper()``` function [@Oksanen2016].

@@

Three of our five treatment years fell in years of below average rainfall (Fig. 2A).
Thus, those three years represent a lower magnitude of absolute change in precipitation experienced by the treatments.
ANPP varied from a minimum of 74.5 g m$^{-2}$ in 2014 to a maximum of 237.1 g m$^{-2}$ in 2016 when averaged across treatments (Fig. 2C).
ANPP was slightly higher in irrigation plots and slightly lower in drought plots (Fig. 2C), corresponding to estimated soil volumetric water content (VWC) differences among treatments (Fig. 2B).
Such differences in soil VWC indicate our treatment infrastructure was successful.
ANPP was highly variable within years (Fig. 2C).

@@

Community composition was similar among treatments.
In no year did community composition among treatments not overlap, and they were equally dispersed in all years (Table 2; Fig. 5).
Community composition was also remarkably stable over time, with no evidence of divergence among treatments (Table 2; Fig. 5).

@@

Ecosystem response to precipitation extremes depends on the physiological responses of constituent species and the rate at which community composition shifts to favor species better able to take advantage of, or cope with, new resource levels [@Smith2009].
Previous work has shown that community compositional shifts can be both rapid, on the order of years [@Hoover2014], and slow, on order of decades [@Knapp2012; @Wilcox2016].
Thus, a lingering question is how the time scales of ecosystem response and community change differ among ecosystems, which can be answered by manipulating precipitation to reach extreme levels.

The results of our five year experiment in a sagebrush steppe conform to two of our four predictions, depending on treatment.
Neither ecosystem functional response nor community composition changed under chronic drought (Fig. 3A, Fig. 4), consistent with the top left prediction in Fig. 1.
Ecosystem functional response under chronic irrigation was different from the control treatment, but community composition remained unchanged (Fig. 3A, Fig. 4), consistent with the top right prediction in Fig. 1.
The altered ecosystem functional response under irrigation matched our expectations because we found evidence for a saturating response for the irrigation treatment (Fig. 3A, right panel).
However, the response of ANPP to soil moisture in all treatments was consistently weak (Table 1, Fig. A2-2) and within-year variation of ANPP was large (Fig. 3B)

@@

First, it could be that our drought manipulation was not large enough to induce a response.
That is, a 50\% decrease in any given year may not be abnormal given our site's historical range of variability [@Knapp2017].
We cannot definitively rule out this possibility, but we have reason to believe our drought treatment *should* have been large enough.
Using the methods described by @Lemoine2016, we calculated the percent reduction and increase of mean growing season precipitation necessary to reach the 1\% and 99\% extremes of the historical precipitation regime at our site (Fig. A3-3).
The 1\% quantile of precipitation at our site is 110 mm, a 47% reduction from the mean, and the 99\% quantile is 414 mm, a 77\% increase from mean growing season precipitation (Appendix 3).
Thus, our drought treatment represented extreme precipitation amounts, especially in years where ambient precipitation was below average (Fig. 2A).
The irrigation treatment may have been too small, yet that is the treatment where we did observe an effect (Fig. 3).

>>>>

---
title: Ecosystem functional response across precipitation extremes in a sagebrush steppe
preprint: false
author: 
  - name: Andrew T. Tredennick
    affiliation: 1
    corresponding: true
    email: atredenn@gmail.com
  - name: Andrew R. Kleinhesselink
    affiliation: 1,2
  - name: Bret Taylor
    affiliation: 3
  - name: Peter B. Adler
    affiliation: 1
affiliation:
  - code: 1
    address: Department of Wildland Resources and the Ecology Center, Utah State University, Logan, Utah 84322
  - code: 2
    address: Department of Ecology and Evolutionary Biology, University of California, Los Angeles, Los Angeles, California 90095
  - code: 3
    address: United States Department of Agriculture, Agriculture Research Station, U.S. Sheep Experiment Station, Dubois, Idaho 83423
abstract: |
  **Background.** Precipitation is predicted to become more variable in the western United States, meaning years of above and below average precipitation will become more common. Periods of extreme precipitation are major drivers of interannual variability in ecosystem functioning in water limited communities, but how ecosystems respond to these extremes over the long-term may shift with precipitation means and variances. Long-term changes in ecosystem functional response could reflect compensatory changes in species composition or species reaching physiological thresholds at extreme precipitation levels.

  **Methods.** We conducted a five year precipitation manipulation experiment in a sagebrush steppe ecosystem in Idaho, United States. We used drought and irrigation treatments (approximately 50% decrease/increase) to investigate whether ecosystem functional response remains consistent under sustained high or low precipitation. We recorded data on aboveground net primary productivity (ANPP), species abundance, and soil moisture. We fit a generalized linear mixed effects model to determine if the relationship between ANPP and soil moisture differed among treatments. We used nonmetric multidimensional scaling to quantify community composition over the five years.

  **Results.** Ecosystem functional response, defined as the relationship between soil moisture and ANPP was similar among drought and control treatments, but the irrigation treatment had a lower slope than the control treatment. However, ANPP response to availabile soil moisture was weak and uncertain regardless of treatment, with all slopes overlapping zero. There was also large spatial variation in ANPP within-years. Plant community composition was remarkably stable over the course of the experiment and did not differ among treatments.

  **Discussion.** Despite some evidence that ecosystem functional response became less sensitive under sustained wet conditions, the response of ANPP to soil moisture was consistently weak and community composition was stable. Differences in ecosystem functional responses across treatments were not related to compensatory shifts at the plant community level, but instead may reflect the insensitivity of the dominant species to soil moisture. These species may be successful precisely because they have evolved life history strategies which buffer them against precipitation variability.
header-includes: >
  \usepackage{setspace}
  \usepackage{todonotes}
  \usepackage{rotating}
  \usepackage{color, soul}
  \usepackage{sectsty}
  \usepackage{bm,mathrsfs}
bibliography: ~./usses_water.bib
output: 
  rticles::peerj_article:
    number_sections: false
    base_format: rmarkdown::pdf_document # bookdown::pdf_document2 # for using \@ref()
---

@@

At any given site, the relationship between aboveground net primary productivity (ANPP) and water availability (e.g., soil moisture) can be characterized by regressing historical observations of ANPP on observations of soil moisture. The fitted functional response can then be used to project ANPP under future precipitation regimes (give example REF). 
A problem with this approach is that it requires extrapolation if future precipitation falls outside outside the historical range of variability [@Smith2011].
For example, the soil moisture-ANPP relationship may be linear within the historical range of interannual variation, but could saturate at higher levels of soil moisture. In fact, saturating relationships are actually common [@Hsu2012; @Gherardi2015a], perhaps because other resources, like nitrogen, become more limiting in wet years than dry years.
Failure to accurately estimate the curvature of the soil moisture-ANPP relationship will lead to over- or underprediction of ANNP under extreme precipitation.

Another problem with relying on historical ecosystem functional responses to predict impacts of altered precipitation regimes is that these relationships themselves might shift over the long-term.
Changes in species identities and abundances can alter an ecosystem's functional response to water availability because different species have different physiological thresholds.
@Smith2009 introduced the 'Hierarchical Response Framework' for understanding the interplay of community composition and ecosystem functioning in response to long-term shifts in resources.
In the near term, ecosystem functioning such as ANPP will reflect the physiological responses of individual species to the manipulated resource level.
For example, ANPP may decline under simulated drought because the initial community consisted of drought-intolerant species [@Hoover2014].
Ecosystem functioning may recover over longer time spans as new species colonize or initial species reorder in relative abundance.
It is also possible that ecosystem functioning shifts to a new mean state, reflecting the suite of species in the new community [@Knapp2012].

Experimental manipulations of limiting resources, like precipitation, offer a route to understanding how ecosystems will respond to resource levels that fall outside the historical range of variability [@Avolio2015; @Gherardi2015; @Knapp2017].
Altering the amount of precipitation over many years should provide insight into the time scales at which water-limited ecosystems respond to chronic resource alteration.
We propose four alternative predictions for the effect of precipitation manipulation on the ecosystem functional response to soil moisture based on the Hierarchical Response Framework (Fig. 1).
We define 'ecosystem functional response' as the relationship between available soil moisture and ANPP.
The four predictions are based on possible outcomes at the community (e.g., community composition) and ecosystem (e.g., soil moisture-ANPP regression) levels.

First, altered precipitation might have no effect on either ecosystem functional response or community composition (Fig. 1, top left).
In this case, changes in ANPP would be well predicted by the current, observed soil moisture-ANPP relationship.
This corresponds to the early phases of the Hierarchical Response Framework, where ecosystem response follows the physiological responses of individual species.
Second, the ecosystem functional response might change while community composition remains the same (Fig. 1, top right).
A saturating soil moisture-ANPP response fits this scenario, where individual species hit physiological thresholds or are limited by some other resource.
Third, the ecosystem functional response might be constant but community composition changes (Fig. 1, bottom left).
In this case, changes in species' identities or abundances occur in response to altered precipitation levels and species more suited to the new conditions compensate for reduced function of initial species.
Fourth, and last, both ecosystem functional response and community composition could change (Fig. 1, bottom right).
New species, or newly abundant species, with different physiological responses completely reshape the ecosystem functional response.

All four outcomes are possible in any given ecosystem, but the time scales at which the different scenarios play out likely differ [@Smith2009; @Wilcox2016; @Knapp2017].
To determine these time scales, we need to amass information on how quickly ecosystem functional responses change in different ecosystems.
We also need to understand whether changes at the ecosystem level are driven by community level changes or individual level responses.

To that end, here we report the results of a five-year precipitation manipulation experiment in a sagebrush steppe grassland.
We imposed drought and irrigation treatments (approximately $\pm50\%$) and measured ecosystem (ANPP) and community (species composition) responses.
We focus on how the drought and irrigation treatments affect the relationship between interannual variation in available soil moisture and interannual variation in ANPP, and if community dynamics underlie the ecosystem responses.
In particular, we are interested in the consistency of the soil moisture-ANPP relationship among treatments.
Is the relationship steeper under the drought treatment, at low soil moisture?
Does the relationship saturate under the irrigation treatment, at high soil moisture?
To answer these questions we fit a generalized linear mixed effects model to test whether the regressions differed among treatments.
We also analyzed community composition over time, allowing us to place our experimental results within the framework our competing predictions (Fig. 1).

@@

## Study Area
We conducted our precipitation manipulation experiment in a sagebrush steppe community at the USDA-ARS Sheep Experimental Station (USSES) near Dubois, Idaho (44.2$^{\circ}$ N, 112.1$^{\circ}$ W), 1500 m above sea level.
The plant community is dominated by the shrub *Artemesia tripartita* and three perennial bunchgrasses, *Pseudoroegneria spicata*, *Poa secunda*, and *Hesperostipa comata*.
During the period of our experiment (2011 -- 2015), average mean annual precipitation was 265 mm year$\phantom{}^{-1}$ and mean monthly temperature ranged from -5.2$^{\circ}$C in January to 21.8$^{\circ}$C in July.
Between 1926 and 1932, range scientists at the USSES established 26 permanent 1 m$^2$ quadrats to track vegetation change over time.
In 2007, we relocated 14 of the original quadrats, six of which were inside a large, permanent livestock exclosure.
We use these six plots as control plots (i.e. ambient precipitation) in the experiment described below.

## Precipitation Experiment
In spring 2011, we established 16 new 1 m$^2$ plots located in the same exclosure as the six control plots.
We avoided areas on steep hill slopes, areas with greater than 20\% cover of bare rock, and areas with greater than 10\% cover of the shrubs *Purshia tridentata* and/or *Amelanchier utahensis*.
We established the new plots in pairs and randomly assigned each plot in a pair to receive a "drought" or "irrigation" treatment.

Drought and irrigation treatments were designed to decrease and increase the amount of ambient precipitation by 50\%.
To achieve this, we used a system of rain-out shelters and automatic irrigation [@Gherardi2013].
The rain-out shelters consisted of transparent acrylic shingles 1-1.5 m above the ground that covered an area of $2.5\times2$ m.
The shingles intercepted approximately 50\% of incoming rainfall, which was channeled into 75 liter containers.
Captured rainfall was then pumped out of the containers and sprayed on to the adjacent irrigation plot via two suspended sprinklers.
Pumping was triggered by float switches once water levels reached about 20 liters.
We disconnected the irrigation pumps each October and reconnected them each April.
The rain-out shelters remained in place throughout the year.

We monitored soil moisture in four of the drought-irrigation pairs using Decagon Devices (Pullman, Washington) 5TM and EC-5 soil moisture sensors.
We installed four sensors around the edges of each 1x1 m census plot, two at 5 cm soil depth and two at 25 cm soil depth.
We also installed four sensors in areas nearby the four selected plot pairs to measure ambient soil moisture at the same depths.
Soil moisture measurements were automatically logged every four hours.
We coupled this temporally intensive soil moisture sampling with spatially extensive readings taken at six points within all 16 plots and associated ambient measurement areas.
These snapshot data were collected on 06-06-2012, 04-29-2015, 05-07-2015, 06-09-2015, and 05-10-2016\footnote{Dates formatted as: mm-dd-yyyy.} using a handheld EC-5 sensor.

Analyzing the response to experimental treatments was complicated by the fact that we did not directly monitor soil moisture in each plot on each day of the experiment.
Only a subset of plots were equipped with soil moisture sensors, and within those plots, one or more of the sensors frequently failed to collect data.
To remedy these problems, and to produce average daily soil moisture values for the ambient, drought, and irrigation treatments, we used a statistical model to predict the average treatment effects on soil moisture during the course of the experiment.

@@

We then modeled the daily deviation from ambient conditions of the drought and irrigation treatments using a linear mixed effects model with independent variables for treatment, season (winter, spring, summer, fall), rainfall, and all two-way interactions. 
Rainy days were defined as any day in which precipitation was recorded and average temperature was above 3$^{\circ}$C.
The day immediately following rainfall was also classified as rainy.
We fit the model using the ```lme4::lmer()``` function [@Bates2015] in R [@R2016], with random effects for plot group and date.
We weighted observations by the number of unique sensors or spot measurements that were taken in each plot on that day.
We then used the model to predict the average daily soil moisture in the treated plots based on the average daily ambient soil moisture.
We could only predict soil moisture in the treated plots on days for which we took at least one ambient soil moisture measurement.
<!--- If we aren't including the code and results for this model, you should
include a reference, probably to Andy's dissertation--->

## Data Collection
We estimated aboveground net primary productivity (ANPP) using a radiometer to relate ground reflectance to plant biomass [see @Byrne2011 for a review].
We recorded ground reflectance at four wavelengths, two associated with red reflectance (626 nm and 652 nm) and two associated with near-infrared reflectance (875 nm and 859 nm).
At each plot in each year, we took four readings of ground reflectances at the above wavelengths.
We also took readings in 12 (2015), 15 (2012, 2013, 2014), or 16 (2016) calibration plots adjacent to the experimental site, in which we harvested all aboveground biomass produced in the current year (we excluded litter and standing dead material), dried it to a constant weight at 60$^{\circ}$C, and weighed it to estimate ANPP. We harvested at peak biomass each year, typically in late June.

For each plot and year, we averaged the four readings for each wavelength and then calculated a greenness index based on the same bands used to calculate NDVI using the MODIS and AVHRR [YOU HAVEN'T DEFINED THESE] bands for NDVI.
We regressed the greenness index against the dry biomass weight from the ten calibration plots to convert the greenness index to ANPP.
We fit regressions to a MODIS-based index and an AVHRR-based index for each year and retained the regression with the better fit based on $R^2$ values.
We then predicted ANPP using the best regression equation for each year (Appendix 1).

Species composition data came from two sources: yearly census maps for each plot made using a pantograph [@Hill1920] and yearly counts of annual species in each plot.
From these sources, we determined the density of all annuals and perennials forbs, the basal cover of perennial grasses, and the canopy cover of shrubs.
We made a large plot-treatment-year by species matrix, where columns were filled with either basal cover or density, depending on the measurement made for the particular species.
We standardized the values in each column so we could directly compare species quantified with different metrics (density, basal cover, and canopy cover).
This puts all abundance values on the same scale, meaning that common and rare species are weighted equally.
Assuming that rare species will respond to treatments more than common ones, our approach is biased towards detecting compositional changes.

@@

\noindent{}where $\mu_{i}$ is the deterministic prediction from the regression model for observation *i*, which is associated with plot *j* and treatment year *t*. 
$\boldsymbol{\beta}$ is the vector of coefficients for the fixed effects in the design matrix $\textbf{X}$. 
Each row of the design matrix represents a single observation ($\textbf{x}_i$) and is a vector with the following elements: 1 for the intercept, a binary 0 or 1 if the treatment is "drought", a binary 0 or 1 if the treatment is "irrigation", the scaled value of VWC, binary "drought" value times VWC, and binary "irrigation" value times VWC. 
Thus, our model treats "control" observations as the main treatment and then estimates intercept and slope offsets for the "drought" and "irrigation" treatments. 
We use our model to test the following hypotheses:

@@

These hypotheses are based on evidence that precipitation-ANPP relationships often saturate with increasing precipitation (REFERENCES).
<!---these should probably be stated in the Intro--->

@@

\noindent{}where $\Sigma$ is the variance-covariance matrix and *j(i)* reads as "plot *j* associated with observation *i*". 
The random year effects ($\boldsymbol{\eta}$) are drawn from a normal prior with mean 0 and standard deviation $\sigma_{\text{year}}$, which was drawn from a half-Cauchy distribution. 
A full description of our model is in Appendix 2.
<!---call out available computer code here?--->

@@

We used nonmetric multidimensional scaling (NMDS) based on Bray-Curtis distances to identify temporal changes in community composition among treatments.
We first calculated Bray-Curtis distances among all plots for each year of the experiment and then extracted those distances for use in the NMDS.
Some values of standardized species' abundances were negative, which is not allowed for calculating Bray-Curtis distances.
We simply added '2' to each abundance value to ensure all values were greater than zero.
We plotted the first two axes of NMDS scores to see if community composition overlapped, or not, among treatments in each year.
We used the ```vegan::metaMDS()``` function [@Oksanen2016] to calculate Bray-Curtis distances and then to run the NMDS analysis.
We used the ```vegan::adonis()``` function [@Oksanen2016] to perform permutational multivariate analysis of variance to test whether treatment plots formed distinct groupings.
To test whether treatment plots were equally dispersed, or not, we used the ```vegan::betadisper()``` function [@Oksanen2016].

@@

Three of our five study years received below average precipitation (Fig. 2A).
Because our treatments are proportional changes in ambient precipitation, those three years represent a lower magnitude of absolute difference in precipitation among the treatments.
<!---why is the previous point important here? should it go in Methods?--->
ANPP varied from a minimum of 74.5 g m$^{-2}$ in 2014 to a maximum of 237.1 g m$^{-2}$ in 2016 when averaged across treatments (Fig. 2C).
ANPP was slightly higher in irrigation plots and slightly lower in drought plots (Fig. 2C), corresponding to estimated soil volumetric water content (VWC) differences among treatments (Fig. 2B).
Such differences in soil VWC indicate our treatment infrastructure was successful.
ANPP was highly variable across plots within years (Fig. 2C).

@@

Community composition was similar among treatments.
Community composition among treatments overlapped in all years and was equally dispersed in all years (Table 2; Fig. 5).
Community composition was also remarkably stable over time, with no evidence of divergence among treatments (Table 2; Fig. 5).

@@

Ecosystem response to precipitation extremes depends on the physiological responses of constituent species and the rate at which community composition shifts to favor species better able to take advantage of, or cope with, new resource levels [@Smith2009].
Previous work has shown that community compositional shifts can be both rapid, on the order of years [@Hoover2014], and slow, on order of decades [@Knapp2012; @Wilcox2016].
A lingering question is how the time scales of ecosystem response and community change vary among ecosystems. Precipitation manipulation experiments can help answer this question, especially if they push water availability outside the historical range of variability for long periods.

The results of our five year experiment in a sagebrush steppe conform to two of our four predictions, depending on treatment.
Neither ecosystem functional response nor community composition changed under chronic drought (Fig. 3A, Fig. 4), representin the top left SCENARIO in Fig. 1.
Ecosystem functional response under chronic irrigation was different from the control treatment, but community composition remained unchanged (Fig. 3A, Fig. 4), representing the top right SCENARIO in Fig. 1.
The altered ecosystem functional response under irrigation matched our expectations because we found evidence for a saturating response for the irrigation treatment (Fig. 3).
<!--- Can we really call it saturating if it is concave up? More accurate to say weaker response than control? --->
However, the response of ANPP to soil moisture in all treatments was consistently weak (Table 1, Fig. A2-2) and within-year variation of ANPP was large (Fig. 3B).
<!---Last sentence leaves me hanging...what's your point? Did it the response change or not? Statistically but not ecologically significant?--->

@@

First, it could be that our drought manipulation was not large enough to induce a response.
That is, a 50\% decrease in any given year may not be abnormal given our site's historical range of variability [@Knapp2017].
We cannot definitively rule out this possibility, but we have reason to believe our drought treatment *should* have been large enough.
Using the methods described by @Lemoine2016, we calculated the percent reduction and increase of mean growing season precipitation necessary to reach the 1\% and 99\% extremes of the historical precipitation regime at our site (Fig. A3-3).
The 1\% quantile of precipitation at our site is 110 mm, a 47% reduction from the mean, and the 99\% quantile is 414 mm, a 77\% increase from mean growing season precipitation (Appendix 3).
Thus, our drought treatment represented extreme precipitation amounts, especially in years where ambient precipitation was below average (Fig. 2A).
The irrigation treatment may not qualify as extreme, yet that is the treatment where we did observe an effect (Fig. 3).
